---
title: "Practice Lecture 18 MATH 342W Queens College"
author: "Professor Adam Kapelner"
---

# Joins

Another one of the core data munging skills is joining data frames together. In the real world, databases consist of multiple dataframes called "tables" and design matrices are built by gathering data from among many tables. To illustrate this, we load two datasets from the package `nycflights13`, one dataset about weather and one about airports:

```{r}
pacman::p_load(nycflights13, data.table, tidyverse, magrittr)
data(weather)
summary(weather)
data(airports)
summary(airports)

head(weather)
head(airports)
```

Note how the weather and airports datasets contain a common feature: name of airport. It is called `FAA` in airports and `origin` in weather.

First we rename the column in weather to match the column in airports:

```{r}
weather %<>% 
  rename(faa = origin)
```

We also pare down the datasets so we can see the joins more clearly:

```{r}
airports %<>% 
  select(faa, lat, lon)
weather %<>% 
  select(faa, time_hour, temp, humid, wind_speed, pressure, wind_gust)
head(airports)
head(weather)
airports_dt = data.table(airports)
weather_dt = data.table(weather)
```

Some features just aren't measured that often e.g. `wind_gust`.

Let's do some joins. First "left". This is likely the most common because it's usually how we conceptualize what we're doing in our heads. Tidyverse has it's own syntax and data.table uses the base R syntax directly. However, it offers a faster syntax (see benchmark below).

```{r}
airports_and_weather = left_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)

airports_and_weather_dt = merge(airports_dt, weather_dt, by = "faa", all.x = TRUE)
airports_and_weather_dt = merge(airports_dt, weather_dt, all.x = TRUE) #note this works too since it knows faa is the only column in common but not recommended since specifying "by" is more clear
airports_and_weather_dt[sample(1 : .N, 500)]
```

Now "right" join.

```{r}
airports_and_weather = right_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
airports_and_weather_dt = merge(airports_dt, weather_dt, by = "faa", all.y = TRUE)
airports_and_weather_dt = merge(airports_dt, weather_dt, all.y = TRUE)
airports_and_weather_dt[sample(1 : .N, 500)]
```


```{r}
airports_and_weather = inner_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
airports_and_weather_dt = merge(airports_dt, weather_dt, by = "faa")
airports_and_weather_dt = merge(airports_dt, weather_dt)
airports_and_weather_dt[sample(1 : .N, 500)]
```

And full, keeping all the rows. We use a subset to show how this works:

```{r}
airports_without_EWR = airports %>%
  filter(faa != "EWR")
airports_without_EWR_dt = data.table(airports_without_EWR)
airports_without_EWR_and_weather = full_join(airports_without_EWR, weather, by = "faa")
airports_without_EWR_and_weather %>% sample_n(500)
airports_without_EWR_and_weather_dt = merge(airports_without_EWR_dt, weather_dt, by = "faa", all = TRUE)
airports_without_EWR_and_weather_dt = merge(airports_without_EWR_dt, weather_dt, all = TRUE)
airports_without_EWR_and_weather_dt[sample(.N, 500)]
```

Who's faster?

```{r}
microbenchmark(
  left_join_tidy = left_join(airports, weather, by = "faa"),
  left_join_dt = merge(airports_dt, weather_dt, by = "faa", all.x = TRUE),
  left_join_R = merge(data.frame(airports_dt), data.frame(weather_dt), by = "faa", all.x = TRUE),
  full_join_tidy = inner_join(airports, weather, by = "faa"),
  full_join_dt = merge(airports_dt, weather_dt, by = "faa"),  
  full_join_R = merge(data.frame(airports_dt), data.frame(weather_dt), by = "faa"), 
  times = 10
)
```

This is one case where tidyverse is actually faster! Base R is as usual, terrible. 

But if we key up the join columns and preorder the datasets...

```{r}
airports = airports %>% arrange(faa)
weather = weather %>% arrange(faa)

setorder(airports_dt, "faa")
setorder(weather_dt, "faa")
setkey(airports_dt, "faa")
setkey(weather_dt, "faa")
microbenchmark(
  left_join_tidy = left_join(airports, weather, by = "faa"),
  left_join_dt = merge(airports_dt, weather_dt, by = "faa", all.x = TRUE),
  full_join_tidy = inner_join(airports, weather, by = "faa"),
  full_join_dt = merge(airports_dt, weather_dt, by = "faa"),
  times = 500
)
```

Basically comparable now. But I did some more benchmarks with larger datasets and I think data.table usually wins (but not by much). 

There is also `semi_join` and `anti_join` that do the opposite of joining. In my experience, these use cases are limited so we'll just not cover them in this class.


