---
title: "Lab 11"
author: "Your Name Here"
output: pdf_document
---

# Asymmetric Cost Modeling, ROC and DET curves

Source the code from this unit in the previous lab.

Plot a DET curve oos and interpret.

```{r}
#TO-DO
```

#TO-DO interpretation


# Missing Data

Load up the Boston Housing Data and separate into matrix `X` for the features and vector `y` for the response. Randomize the rows

```{r}
rm(list = ls())
set.seed(1)
boston = MASS::Boston
boston_shuffled = boston[sample(1 : nrow(boston)), ]
X = as.matrix(boston_shuffled[, 1 : 13])
y = boston_shuffled$medv
rm(boston, boston_shuffled)
```


Similar to lab 1, write a function that takes a matrix and punches holes (i.e. sets entries equal to `NA`) randomly with an argument `prob_missing`.

```{r}
punch_holes = function(mat, prob_missing){
  n = nrow(mat) * ncol(mat)
  is_missing = as.logical(rbinom(n, 1, prob_missing))
  mat[is_missing] = NA
  mat
}
```

Create a matrix `Xmiss` which is `X` but has missingness with probability of 10% using the function you just wrote. 

```{r}
#TO-DO
```

Also, generate the M matrix and delete columns that have no missingness.

```{r}
M = apply(is.na(Xmiss), 2, as.numeric)
colnames(M) = paste("is_missing_", colnames(X), sep = "")
M = M[, colSums(M) > 0]
```

Split the first 400 observations were the training data and the remaining observations are the test set. For Xmiss, cbind on the M so the model has a chance to fit on "is missing" as we discussed in class.

```{r}
train_idx = 1 : 400
test_idx = setdiff(1 : nrow(X), train_idx)
X_train =     X[train_idx, ]
Xmiss_train = cbind(Xmiss, M)[train_idx, ]
y_train =     y[train_idx]
X_test =      X[test_idx, ]
Xmiss_test =  cbind(Xmiss, M)[test_idx, ]
y_test =      y[test_idx]
```

Fit a random forest model of `y_train ~ X_train`, report oos s_e (not oob) on `X_test`. This ignores missingness

```{r}
#TO-DO
sqrt(mean((y_hat_test - y_test)^2))
```

Impute the missingness in `Xmiss` using the feature averages to create a matrix `Ximp_naive_train` and `Ximp_naive_test`. 

```{r}
#TO-DO
```

Fit a random forest model of `y_train ~ Ximp_naive_train`, report oos s_e (not oob) on `Ximp_naive_test`.

```{r}
#TO-DO
sqrt(mean((y_hat_test - y_test)^2))
```

How much predictive performance was lost due to missingness when naive imputation was used vs when there was no missingness?

```{r}
#TO-DO
```

Use `missForest` to impute the missing entries to create a matrix `Ximp_MF_train` and `Ximp_MF_test`.

```{r}
pacman::p_load(missForest)
#TO-DO
```

Fit a random forest model of `y_train ~ Ximp_MF_train`, report oos s_e (not oob) on `Ximp_MF_test`.

```{r}
#TO-DO
sqrt(mean((y_hat_test - y_test)^2))
```

How much predictive performance was lost due to missingness when `missForest` imputation was used?

```{r}
#TO-DO
```

Why did `missForest` imputation perform better than naive imputation?

#TO-DO

Reload the feature matrix:

```{r}
rm(list = ls())
X = as.matrix(MASS::Boston[, 1 : 13])
```


#Boosting

We will make use of YARF's tree-fitting method so here's the boilerplate code to load it once again:

```{r}
options(java.parameters = "-Xmx8000m")
pacman::p_load(rJava)
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
pacman::p_load(YARF)
```

We will now write a gradient boosting algorithm from scratch. We will make it as general as possible for regression and classification.

```{r}
pacman::p_load(checkmate) #this is a package that enforces arguments are the correct form

#' Gradient boosting
#'
#' Generates a gradient boosting model based on your choices of base learner and objective function
#' 
#' @param X                         A data frame representing the features. It is of size n x p. No need for an intercept column.
#' @param y                         A vector of length n. It either will be real numbers (for regression) or binary (for classification).
#' @param g_base_learner_alg        A function with arguments X, y and ... and returns a function that takes X as an argument. The default is YARFCART
#'                                  with nodesize 10% of the total length.
#' @param neg_grad_objective_function   The negative gradient of the function to be minimized. It takes arguments y, yhat that returns a vector. The default objective function is SSE for
#'                                  regression and logistic loss for classification.
#' @param M                         The number of base learners to be summed. Default is 50 for regression and 100 for classification.
#' @param eta                       The step size in the gradient descent. Default is 0.3
#' @param verbose                   Messages are printed out during construction. Default is TRUE.
#' @param ...                       Optional arguments to be passed into the g_base_learner_alg function.
#'
#' @return                          A "qc_basement_gbm" gradient boosting model which can be used for prediction
qc_basement_gbm = function(X, y, g_base_learner_alg = NULL, neg_grad_objective_function = NULL, M = NULL, eta = 0.3, verbose = TRUE, ...){
  assert_data_frame(X)
  n = nrow(X)
  assert_numeric(y)
  assert(length(y) == n)
  assert_function(g_base_learner_alg, args = c("X", "y"), null.ok = TRUE)
  if (is.null(g_base_learner_alg)){
    g_base_learner_alg = function(X0, y0){
      #we want some bias in the base learner - so default to 10% of the sample size
      YARFCART(X0, y0, nodesize = round(.1 * nrow(X0)), calculate_oob_error = FALSE, bootstrap_indices = list(1 : nrow(X0)), verbose = FALSE)
    }
  }
  assert_function(neg_grad_objective_function, args = c("y", "yhat"), null.ok = TRUE)
  assert_count(M, positive = TRUE, null.ok = TRUE)
  assert_numeric(eta, lower = .Machine$double.eps)
  assert_logical(verbose)
  
  g_0 = function(X_star){
    #TO-DO
  }
  if (identical(sort(names(table(y))), c("0", "1"))){
    model_type = "probability_estimation"
    if (verbose){cat("building gradient boosted model for probability estimation of two classes\n")}
    if (is.null(M)){
      M = 100
    }
    if (is.null(neg_grad_objective_function)){
      neg_grad_objective_function = function(y, y_hat){
        #TO-DO
      }
    }
  } else {
    model_type = "regression"
    if (verbose){cat("building gradient boosted model for regression\n")}
    if (is.null(M)){
      M = 50
    }
    if (is.null(neg_grad_objective_function)){
      neg_grad_objective_function = function(y, y_hat){
        #TO-DO
      }
    }
  }

  #these are the partial function fits where G_{t+1} = G_t + eta * gtilde_t
  g_tildes = list() 
  #this is the running tally of the sum of the yhats at each iteration
  cumul_y_hat_m = #TO-DO 
  for (m in 1 : M) {
    if (verbose){cat("fitting base learner", m, "of", M, "\n")}
    g_tildes[[m]] = #TO-DO
    cumul_y_hat_m = cumul_y_hat_m + #TO-DO
  }
  
  gbm = list(
    M = M, 
    eta = eta,
    X = X, 
    y = y, 
    model_type = model_type,
    neg_grad_objective_function = neg_grad_objective_function, 
    g_base_learner_alg = g_base_learner_alg,
    g_0 = g_0,
    g_tildes = g_tildes
  )
}

#' Compute all iterative boosting predictions
#' 
#' Returns all predictions for each iteration of the gradient boosting
#'
#' @param gbm     A gradient boosting model of class "qc_basement_gbm"
#' @param X_star  The data to predict for (as a data frame). It has n_* rows and p columns
#'
#' @return        A matrix with n_* rows and M+1 columns where each column are the iterative
#'                predictions across all base learners beginning with g_0. For regression, the
#'                unit is in the units of the original response. For probability estimation for 
#'                binary response, the unit is the logit of the probability estimate.
qc_basement_gbm_all_predictions = function(gbm, X_star){
  assert_class(gbm, "qc_basement_gbm")
  assert_data_frame(X_star)
  
  all_y_hat_star = matrix(NA, nrow = nrow(X_star), ncol = gbm$M + 1)
  all_y_hat_star[, 1] = #TO-DO
  for (m in 1 : gbm$M){
    all_y_hat_star[, m + 1] = #TO-DO
  } 
  all_y_hat_star
}


#' GBM Predict
#' 
#' Returns final predictions for the gradient boosting model
#'
#' @param gbm     A gradient boosting model of class "qc_basement_gbm"
#' @param X_star  The data to predict for (as a data frame). It has n_* rows and p columns
#'
#' @return        A vector of length n_* rows with each row's predictions. For regression, the
#'                unit is in the units of the original response. For probability estimation for 
#'                binary response, the unit is the logit of the probability estimate.
qc_basement_gbm_predict = function(gbm, X_star){
  #TO-DO
}
```

Now we test the code in-sample:

```{r}
set.seed(1)
n = 100
p = 3
X = matrix(rnorm(n * p), nrow = n)
bbeta = seq(-1, 1, length.out = p)
y = c(X %*% bbeta + rnorm(n))
y_binary = rbinom(n, 1, 1 / (1 + exp(-X %*% bbeta)))
X = data.frame(X)

#regression
g_b = qc_basement_gbm(X, y)
pacman::p_load(ggplot2)
ggplot(data.frame(y = y, yhat = qc_basement_gbm_predict(g_b, X))) + aes(x = y, y = yhat) + geom_point()
y_hats_by_m = qc_basement_gbm_all_predictions(g_b, X)
rmses_by_m = apply(y_hats_by_m, 2, function(y_hat){sqrt(mean((y - y_hat)^2))})
rmses_by_m

#probability estimation
g_b = qc_basement_gbm(X, y_binary)
table(y_binary, as.numeric(qc_basement_gbm_predict(g_b, X) > 0))
y_hats_by_m = qc_basement_gbm_all_predictions(g_b, X) > 0
miscl_err_by_m = apply(y_hats_by_m, 2, function(y_hat){mean(y_binary != y_hat)})
miscl_err_by_m
```


Here is code to split up the diamonds dataset into three subsets:

```{r}
set.seed(1)
diamonds = ggplot2::diamonds
pacman::p_load(tidyverse)
diamonds = diamonds %>% 
  mutate(cut = factor(cut, ordered = FALSE)) %>%
  mutate(color = factor(color, ordered = FALSE)) %>%
  mutate(clarity = factor(clarity, ordered = FALSE))
diamonds_mm = model.matrix(price ~ ., diamonds)
train_size = 2000
train_indices = sample(1 : nrow(diamonds), train_size)

y_train = diamonds[train_indices, ]$price
X_train = diamonds_mm[train_indices, ]

validation_size = 2000
validation_indices = sample(setdiff(1 : nrow(diamonds), train_indices), validation_size)
y_validation = diamonds[validation_indices, ]$price
X_validation_mm = diamonds_mm[validation_indices, ]

test_size = 2000
test_indices = sample(setdiff(1 : nrow(diamonds), c(train_indices, validation_indices)), test_size)
y_test = diamonds[test_indices, ]$price
X_test_mm = diamonds_mm[test_indices, ]
```

Using your new gradient boosting function, optimize the number of base learners, M for the diamonds data using a grid search:

```{r}
#TO-DO
```

Now find the error in the test set and comment on its performance:

```{r}
#TO-DO
```

Repeat this exercise for the adult dataset. First create the splits:

```{r}
#TO-DO
```

Using your new gradient boosting function, optimize the number of base learners, M for the diamonds data using a grid search:

```{r}
#TO-DO
```

Now find the error in the test set and comment on its performance:

```{r}
#TO-DO
```


